Elaboration with First-Class Implicit Function Types

### 2020-02-29 paper.tex
stuff

In the definition of checking, we insert a new clause after
$\echeck{t}{\Theta}{\Gamma}{(\{x : A\}\to B)}$:
\begin{alignat*}{3}
& \echeck{t}{\Theta_0}{\Gamma}{(\alpha\,\ol{t})} :\equiv \edo\\
& \quad \slet\,(\Theta_1,\,\theta_1,\,A) = \freshMeta{\Theta_0}{\Gamma}{\Tel}\\
& \quad (\Theta_2,\,\theta_2,\,t',\,B) \leftarrow \einfer{t}{\true}{\Theta_1}{\Gamma[\theta_1],\,x:\Rec\,A}\\
& \quad \slet\,(\Theta_3,\,\theta_3) = \newConstancy{\Theta_2}{\Gamma[\theta_{12}]}{x}{(A[\theta_2])}{B}\\
& \quad (\Theta_4,\,\theta_4) \leftarrow
            \unify\, ((\alpha\,\ol{t})[\theta_{123}])\,(\{x : \ol{A[\theta_{23}]}\} \to B[\theta_3])\\
& \quad \ereturn\,(\Theta_4,\,\theta_{1234},\,(\lambda\,\{x : \ol{A[\theta_{234}]}\}.\,t'[\theta_{34}]))
\end{alignat*}
Hence, when checking a term with a meta-headed type, we create a fresh meta with
$\Tel$ type, infer a type for the term, and wrap the result in a strictly
curried $\lambda$. We again need to create new constancy constraint. This way,
if $B$ does not depend on $x : \Rec\,A$, the strictly curried $\lambda$ in the
output immediately computes away, since $A$ is solved to $\epsilon$.

@@

It is apparent from the $B1$ and $B2$ cases on Figure \ref{fig:benchmark} that FCIF
cannot infer implicit function types for function arguments.  MLF is the only
system which can do this, in cases where the polymorphic argument is used only
once. Consider the following:
\[
  \slet\,f : IdTy \to Bool = \lambda\,\_.\,true\,\sin\,\lambda\,x.\,f\,x
  \]
FCIF will attempt to elaborate $f\,x$ to $f\,(\lambda\{A\}.\,x)$, but fails to
infer a type for $x$, because $A$ is not in the scope of $x$'s type. We sketch
an extension of FCIF with \emph{implicit application insertion}, which could
possibly handle this example. We note though that this is not yet fleshed out
and is a subject of future work.

\section{Future Work and Conclusion}

>>>>

In the definition of checking, we insert a new clause after
$\echeck{t}{\Theta}{\Gamma}{(\{x : A\}\to B)}$:
\begin{alignat*}{3}
& \echeck{t}{\Theta_0}{\Gamma}{(\alpha\,\ol{u})} :\equiv \edo\\
& \quad \slet\,(\Theta_1,\,\theta_1,\,A) = \freshMeta{\Theta_0}{\Gamma}{\Tel}\\
& \quad (\Theta_2,\,\theta_2,\,t',\,B) \leftarrow \einfer{t}{\true}{\Theta_1}{\Gamma[\theta_1],\,x:\Rec\,A}\\
& \quad \slet\,(\Theta_3,\,\theta_3) = \newConstancy{\Theta_2}{\Gamma[\theta_{12}]}{x}{(A[\theta_2])}{B}\\
& \quad (\Theta_4,\,\theta_4) \leftarrow
            \unify\, ((\alpha\,\ol{u})[\theta_{123}])\,(\{x : \ol{A[\theta_{23}]}\} \to B[\theta_3])\\
& \quad \ereturn\,(\Theta_4,\,\theta_{1234},\,(\lambda\,\{x : \ol{A[\theta_{234}]}\}.\,t'[\theta_{34}]))
\end{alignat*}
Hence, when checking a term with a meta-headed type, we create a fresh meta with
$\Tel$ type, infer a type for the term, and wrap the result in a strictly
curried $\lambda$. We again need to create new constancy constraint. This way,
if $B$ does not depend on $x : \Rec\,A$, the strictly curried $\lambda$ in the
output immediately computes away, since $A$ is solved to $\epsilon$.

@@

It is apparent from the $B1$ and $B2$ cases on Figure \ref{fig:benchmark} that FCIF
cannot infer implicit function types for function arguments.  MLF is the only
system which can do this, in cases where the polymorphic argument is used only
once. Consider the following:
\[
  \slet\,f : IdTy \to Bool = \lambda\,\_.\,true\,\sin\,\lambda\,x.\,f\,x
  \]
FCIF will attempt to elaborate $f\,x$ to $f\,(\lambda\{A\}.\,x)$, but fails to
infer a type for $x$, because $A$ is not in the scope of $x$'s type. We sketch
an extension of FCIF with \emph{curried application insertion}, which could
possibly handle this example. We extend $\einsert$ with the following case. We
omit metacontexts and metasubstitutions for brevity.
\begin{alignat*}{3}
& \einsert\,(t,\,\alpha\,\ol{u}) :\equiv \edo \\
& \quad A \leftarrow \mathsf{freshMeta}_{\Gamma}\,\Tel\\
& \quad B \leftarrow \mathsf{freshMeta}_{\Gamma,\,x: \Rec\,A}\,\U\\
& \quad \unify\,(\alpha\,\ol{u})\,(\{x : \ol{A}\} \to B)\\
& \quad \mathsf{newConstancy}_{\Gamma, x : \Rec\,A}\,B\\
& \quad u \leftarrow \mathsf{freshMeta}_{\Gamma}\,(\Rec\,A)\\
& \quad \ereturn\,(t\,\{u : \ol{A}\},\,B[x \mapsto u])
\end{alignat*}
In short, we insert a new curried application whenever we $\einsert$ with
meta-headed type. Now, elaboration returns
$f\,(\lambda\,\{A\}.\,x\,\{\alpha_2\,x\,A : \ol{\alpha_0}\})$, and right before
we try to unify expected and inferred types, we have $x : \{y :
\ol{\alpha_0}\}\to \alpha_1\,y$, and we have to compute
$\unify\,(\alpha_1\,\{\alpha_2\,x\,A : \ol{\alpha_0}\})\,(A \to A)$. This
problem does not fall in any usual pattern fragment, and does not have an
obvious most general solution. However, it is reasonable to make the following
assumption: metas like $\alpha_2$ are only solved with \emph{order-preserving
  embeddings}, so for $\alpha_2$ the four potential solutions are
$(\lambda\,x\,A.\,[])$, $(\lambda\,x\,A.\,(A :: []))$, $(\lambda\,x\,A.\,(x ::
[]))$ and $(\lambda\,x\,A.\,(x :: A :: []))$. The reason for only allowing such
solutions is that these cases can possibly yield solvable pattern unification
problems, while solutions containing arbitrary terms or non-linear variable
occurrences cannot. Out of these, $(\lambda\,x\,A.\,(x :: A :: []))$ and
$(\lambda\,x\,A.\,(x :: []))$ are ruled out because they would require a cyclic
solution for $\alpha_0$, and $(\lambda\,x\,A.\,[])$ is ruled out because it
would yield the unsolvable $\unify\,\alpha_1\,(A \to A)$ problem. Hence, there
is a unique order-preserving embedding solution in this case, which does yield
the expected elaboration output. However, note that this idea is far from being
fully fleshed out, and it could be subject of future research.

\section{Conclusions and Future Work}
